---
title: "Project Proposal Template"
authors: Trieste Cerny and Jackson Justice
format: 
  pdf:
    embed-resources: true 
---

<style type="text/css">
h1.title {
  font-size: 40px;
  text-align: center;
}
</style>

testing git so i can commit it to the correct repo
```{r setup, include=FALSE}
# load all necessary packages here
```

DELETE THIS PARAGRAPH: You should replace all text in this template with the requested content. You will lose 10% of the points on this assignment if the title of your project proposal is still "Project Proposal Template," (provide your own title even if it's just "STAT 330 Project Proposal") or if the text in each section isn't replaced with your own words, **written in complete sentences and paragraphs.** You should also delete the pre-existing comments in the code blocks of this template, providing your own comments where appropriate.

# 1 Group Members and Confirmation That You Have Read Final Project Instructions

DELETE THIS: List the names of all group members in this section. (Don't forget to delete this part! The only content in this section should be a sentence like "The members of our group are...")

DELETE THIS: Confirm that each member of you group has read the **full** 5-page project instructions document Stat_330_Final_Project_Instructions_W26.pdf

# 2 Data Set

DELETE THIS: What dataset do you intend to use for your final project? Where can it be found? **Provide a description of the data set and the variables it contains.** Load your data into R and print the first few rows of the data set.

```{r}
#Read your data into R. 
#You will eventually need to provide a copy of the data to the
#professor as part of the final project submission, so it
#probably makes sense to download a copy of the data to your
#machine rather than call the data from a url.

#proj_data = read.table("PATH_TO_DATA")
```


# 3 Potential Questions of Interest

DELETE THIS: What are two scientific questions you could try to answer by applying regression tools to this dataset? **Provide commentary on why someone might be interested in studying these questions.**

# 4 Exploratory Data Analysis

DELETE THIS: Conduct EDA and provide **at least 3** figures or statistics highlighting features or between variable relationships within the data that are relevant to the questions of interest you provided earlier in the proposal. **You should explain what you found interesting or noteworthy about each of 3 included elements.**
```{r}
#Code for your EDA here. 
#Can you come up with useful visualizations of your data 
#that might help someone gain insight into your questions of interest 
#prior to any formal statistical analysis being conducted?
```

# 5 Preliminary Model Fit

DELETE THIS: Select a variable from your dataset that makes sense to describe as a dependent or response variable, and another to describe as an explanatory or predictor variable.

DELETE THIS: **Write out the model you have chosen** and fit it using R. The SLR model you fit doesn't have to directly relate to one of your research questions, but **you should provide a brief justification of why someone may be interested in the relationship characterized by your initial model.**

```{r}
#lm.fit = lm(response~predictor, data = proj_data)
```

DELETE THIS: Assess whether the regression assumptions are satisfied **and provide interpretation for the estimated model coefficients and associated uncertainty estimates.** 

**Even if the assumptions are not fully satisfied, you should interpret the estimated betas and standard errors/confidence intervals as though the model fit was appropriate in order to demonstrate that you know how to do this.** You are welcome to use as many or as few code blocks as you wish in this section, but make sure that your analysis is easy to follow.

```{r}
#Assess assumptions

#Interpret model output contained in summary(lm.fit)
```

