---
title: "Stat 330 Project Proposal - Jackson and Trieste"
authors: Jackson Justice and Trieste Cerny
format: 
  pdf:
    embed-resources: true 
---

```{=html}
<style type="text/css">
h1.title {
  font-size: 40px;
  text-align: center;
}
</style>
```

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
```

# 1 Group Members and Confirmation That You Have Read Final Project Instructions

The members of our group are Jackson Justice and Trieste Cerny. We both have read the full 5-page project instructions and we understand the expectations for the Stat 330 Final Project.

# 2 Data Set

For our final project, we will use the College Scorecard Institution-Level Dataset. This dataset contains information on over 6,000 higher education institutions in the United States.

Of particular interest to us is the variable MD_EARN_WNE_P10, which represents the median earnings of former students 10 years after entry. This allows us to evaluate long-term financial outcomes associated with different colleges.

```{r}
#Read your data into R. 
proj_data <- read.csv("college-scorecard-institution.csv.gz")
```

# 3 Potential Questions of Interest

Question 1: Does attending a more expensive college lead to higher median earnings 10 years after entry, after controlling for incoming student quality such as average SAT scores?

Many students assume that going to a more prestigious and therefore more expensive university will result in better long-term financial outcomes. However, higher tuition institutions may also enroll students who were already academically stronger. By controlling for incoming student quality, we can attempt to find whether tuition itself is associated with higher earnings outcomes.

Question 2: Do institutions with a higher percentage of students receiving Pell Grants tend to have different median earnings outcomes 10 years after entry?

The percentage of students receiving Pell Grants reflects the socioeconomic composition of a college. Colleges that enroll a higher proportion of low-income students may face different structural challenges or provide different types of support compared to institutions with fewer Pell Grant recipients. By examining whether this percentage is associated with median earnings 10 years after entry, we can better understand how institutional demographics relate to long-term economic outcomes for students.

\# 4 Exploratory Data Analysis

```{r}
# Clean numeric variables
proj_data$NPT4_PUB[proj_data$NPT4_PUB == "NULL"] <- NA
proj_data$NPT4_PRIV[proj_data$NPT4_PRIV == "NULL"] <- NA
proj_data$MD_EARN_WNE_P10[proj_data$MD_EARN_WNE_P10 == "NULL"] <- NA
proj_data$PCTPELL[proj_data$PCTPELL == "NULL"] <- NA

proj_data$NPT4_PUB <- as.numeric(proj_data$NPT4_PUB)
proj_data$NPT4_PRIV <- as.numeric(proj_data$NPT4_PRIV)
proj_data$MD_EARN_WNE_P10 <- as.numeric(proj_data$MD_EARN_WNE_P10)
proj_data$PCTPELL <- as.numeric(proj_data$PCTPELL)

# create one price variable for public and private schools
proj_data$NET_PRICE <- ifelse(!is.na(proj_data$NPT4_PUB),
                              proj_data$NPT4_PUB,
                              proj_data$NPT4_PRIV)

#EDA 1
histogram_of_earnings <- ggplot(proj_data, aes(x = MD_EARN_WNE_P10)) +
  geom_histogram(bins = 50, na.rm = TRUE) +
  labs(title = "Distribution of Median Earnings 10 Years After College Entry",
       x = "Median Earnings",
       y = "Count") +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(labels = scales::comma)
histogram_of_earnings

```

Histogram of Earnings is a distribution of median earnings, and is right-skewed. This suggests that earnings are not evenly distributed across all institutions.

```{r}
#EDA 2
ggplot(proj_data, aes(x = NET_PRICE, y = MD_EARN_WNE_P10)) +
  geom_point(alpha = 0.4, na.rm = TRUE) +
  labs(title = "Price vs Median Earnings",
       x = "Average Tuition Price",
       y = "Median Earnings")+
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(labels = scales::comma)
```

There appears to be a positive relationship between tuition price and median earnings, although the relationship is not perfectly linear. There is substantial variation in earnings among institutions with similar tuition levels.

```{r}
ggplot(proj_data, aes(x = PCTPELL, y = MD_EARN_WNE_P10)) +
  geom_point(alpha = 0.4, na.rm = TRUE) +
  labs(title = "Percent Pell Grant Recipients vs Median Earnings",
       x = "Percent of Students Receiving Pell Grants",
       y = "Median Earnings")+
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(labels = scales::comma)

```

There appears to be a negative relationship between the percentage of Pell Grant recipients and median earnings. Institutions serving a higher proportion of low-income students tend to have lower median earnings outcomes, although the relationship shows variability.

# 5 Preliminary Model Fit

Our response variable is median earnings 10 years after college entry. Our predictor variable is the average net price. Median earnings is treated as the dependent variable because it represents the long-term financial outcome for students, while net price is the explanatory variable since it reflects the cost of attending the institution.

```{r}
lm.fit = lm(MD_EARN_WNE_P10 ~ NET_PRICE, data = proj_data)
summary(lm.fit)
```

Equation: Median Earnings = 30,830 + 0.577 × (Net Price)

The intercept (30,830) represents the predicted median earnings when net price is zero. Although a net price of zero is not realistic, the intercept serves as the baseline of earnings in the model.

The slope coefficient (0.577) indicates that for each additional \$1 increase in average net price, median earnings increase by approximately \$0.58 on average.

The p-value for NET_PRICE is less than 2e-16, which is extremely small. This indicates strong statistical evidence that there is a positive association between net price and median earnings.

The R-squared value is 0.1057. This means that about 10.57% of the variability in median earnings across colleges is explained by net price alone.

There is a statistically significant positive relationship between average net price and median earnings 10 years after college entry. The relatively low R-squared value indicates that net price alone does not strongly predict earning outcomes. Other variables likely play an important role.

```{r}
#residuals vs fitted plot
plot(lm.fit$fitted.values,
     resid(lm.fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red")

#Q-Q plot
qqnorm(resid(lm.fit),
       main = "Normal Q-Q Plot of Residuals")
qqline(resid(lm.fit), col = "red")
```

Linearity: The residuals vs. fitted plot does not show a strong curved pattern, so a linear relationship appears reasonable.

Independence: Each observation represents a different college so they are unlikely to affect one another.

Normality: The Q–Q plot shows noticeable deviation in the tails, indicating that the residuals are not perfectly normally distributed and that there may be some extreme institutions influencing the results.

Equal Variance: The residuals vs. fitted plot shows that the spread of residuals is relatively consistent across predicted values but there is spread at higher predicted earnings.
