---
title: "Stat 330 Project Proposal"
author: ["Jackson Justice", "Trieste Cerny"]
format: 
  pdf:
    embed-resources: true 
---

```{=html}
<style type="text/css">
h1.title {
  font-size: 40px;
  text-align: center;
}
</style>
```

```{r setup, include=FALSE}
library(tidyverse)
```


# 1 Group Members and Confirmation That You Have Read Final Project Instructions

The members of our group are Jackson Justice and Trieste Cerny. We both have read the full 5-page project instructions and we understand the expectations for the Stat 330 Final Project.


# 2 Data Set  

For our final project, we will use the College Scorecard Institution-Level Dataset. The unit of observation is an individual higher education institution in the United States, and the dataset contains information on over 6,000 colleges.

Our primary outcome variable is `MD_EARN_WNE_P10`, which represents the median earnings of former students 10 years after entry. This variable allows us to evaluate long-term financial outcomes associated with different colleges. We will also examine variables related to institutional cost (average net price) and student socioeconomic composition (percentage receiving Pell Grants).

```{r}
#Read your data into R. 
proj_data <- read.csv("college-scorecard-institution.csv")
```


# 3 Potential Questions of Interest

**Question 1:** Is average net price associated with median earnings 10 years after entry, after accounting for measures of incoming student quality such as average SAT scores?

Many students assume that attending a more prestigious and therefore more expensive university results in better long-term financial outcomes. However, institutions with higher tuition may also enroll students who were already academically stronger. By accounting for incoming student quality, we aim to examine whether institutional cost is associated with earnings outcomes beyond differences in student preparedness.

**Question 2:** Is the percentage of students receiving Pell Grants associated with median earnings 10 years after entry?

The percentage of students receiving Pell Grants reflects the socioeconomic composition of an institution. Colleges that enroll a higher proportion of low-income students may differ systematically in selectivity, available resources, or program offerings. Examining this relationship helps us better understand how institutional demographics are related to long-term economic outcomes at the college level.


# 4 Exploratory Data Analysis

```{r}
# Clean numeric variables
proj_data$NPT4_PUB[proj_data$NPT4_PUB == "NULL"] <- NA
proj_data$NPT4_PRIV[proj_data$NPT4_PRIV == "NULL"] <- NA
proj_data$MD_EARN_WNE_P10[proj_data$MD_EARN_WNE_P10 == "NULL"] <- NA
proj_data$PCTPELL[proj_data$PCTPELL == "NULL"] <- NA

proj_data$NPT4_PUB <- as.numeric(proj_data$NPT4_PUB)
proj_data$NPT4_PRIV <- as.numeric(proj_data$NPT4_PRIV)
proj_data$MD_EARN_WNE_P10 <- as.numeric(proj_data$MD_EARN_WNE_P10)
proj_data$PCTPELL <- as.numeric(proj_data$PCTPELL)

# create one price variable for public and private schools
proj_data$NET_PRICE <- ifelse(!is.na(proj_data$NPT4_PUB),
                              proj_data$NPT4_PUB,
                              proj_data$NPT4_PRIV)

#EDA 1
histogram_of_earnings <- ggplot(proj_data, aes(x = MD_EARN_WNE_P10)) +
  geom_histogram(bins = 50, na.rm = TRUE) +
  labs(title = "Distribution of Median Earnings 10 Years After College Entry",
       x = "Median Earnings",
       y = "Count") +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(labels = scales::comma)
histogram_of_earnings

```

The distribution of median earnings 10 years after entry is right-skewed, with most institutions clustered between approximately \$30,000 and \$60,000 and a smaller number of institutions exhibiting substantially higher earnings. This indicates that a relatively small group of colleges have unusually high median earnings compared to the majority.

The skewness and presence of extreme values suggest that influential institutions may affect regression estimates. This observation motivates careful examination of model assumptions and consideration of potential transformations in later stages of the project.

```{r}
#EDA 2
ggplot(proj_data, aes(x = NET_PRICE, y = MD_EARN_WNE_P10)) +
  geom_point(alpha = 0.4, na.rm = TRUE) +
  labs(title = "Price vs Median Earnings",
       x = "Average Tuition Price",
       y = "Median Earnings")+
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(labels = scales::comma)
```

The scatterplot suggests a positive association between average tuition price and median earnings 10 years after entry. Institutions with higher tuition levels tend, on average, to have higher median earnings. However, the relationship is not perfectly linear, and there is considerable vertical spread in earnings at similar tuition levels.

This substantial variability indicates that net price alone does not strongly determine earnings outcomes. Other institutional characteristics, such as selectivity, student composition, or program offerings, likely contribute meaningfully to differences in long-term earnings. The visible spread also suggests that prediction intervals may be wide when modeling earnings using net price alone.

```{r}
ggplot(proj_data, aes(x = PCTPELL, y = MD_EARN_WNE_P10)) +
  geom_point(alpha = 0.4, na.rm = TRUE) +
  labs(title = "Percent Pell Grant Recipients vs Median Earnings",
       x = "Percent of Students Receiving Pell Grants",
       y = "Median Earnings")+
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(labels = scales::comma)

```

The scatterplot indicates a negative association between the percentage of students receiving Pell Grants and median earnings 10 years after entry. Institutions with a higher proportion of Pell Grant recipients tend, on average, to have lower median earnings outcomes.

The downward trend appears relatively consistent across much of the range, although there remains noticeable variability at lower and mid-range Pell percentages. This pattern suggests that institutional socioeconomic composition is strongly related to long-term earnings outcomes, while also indicating that additional factors beyond Pell share contribute to differences in earnings across colleges.


# 5 Preliminary Model Fit

In our preliminary analysis, the response variable is median earnings 10 years after college entry (`MD_EARN_WNE_P10`), and the predictor variable is average net price (`NET_PRICE`). Median earnings is treated as the response because it represents the long-term financial outcome associated with attending an institution, while net price serves as the explanatory variable reflecting institutional cost.

We begin with a simple linear regression model to examine the bivariate association between institutional cost and long-term earnings. This provides an initial framework for understanding the strength and direction of the relationship before extending to more complex models that incorporate additional covariates.

```{r}
lm.fit = lm(MD_EARN_WNE_P10 ~ NET_PRICE, data = proj_data)
summary(lm.fit)
```

Equation: Median Earnings = 30,830 + 0.577 × (Net Price)

The intercept (30,830) represents the predicted median earnings when net price is zero. Although a net price of zero is not realistic, the intercept serves as the baseline of earnings in the model.

The slope coefficient (0.577) indicates that for each additional \$1 increase in average net price, median earnings increase by approximately \$0.58 on average.

The p-value for NET_PRICE is less than 2e-16, which is extremely small. This indicates strong statistical evidence that there is a positive association between net price and median earnings.

The R-squared value is 0.1057. This means that about 10.57% of the variability in median earnings across colleges is explained by net price alone.

There is a statistically significant positive relationship between average net price and median earnings 10 years after college entry. The relatively low R-squared value indicates that net price alone does not strongly predict earning outcomes. Other variables likely play an important role.

```{r}
#residuals vs fitted plot
plot(lm.fit$fitted.values,
     resid(lm.fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red")

#Q-Q plot
qqnorm(resid(lm.fit),
       main = "Normal Q-Q Plot of Residuals")
qqline(resid(lm.fit), col = "red")
```

Linearity: The residuals vs. fitted plot does not show a strong curved pattern, so a linear relationship appears reasonable.

Independence: Each observation represents a different college so they are unlikely to affect one another.

Normality: The Q–Q plot shows noticeable deviation in the tails, indicating that the residuals are not perfectly normally distributed and that there may be some extreme institutions influencing the results.

Equal Variance: The residuals vs. fitted plot shows that the spread of residuals is relatively consistent across predicted values but there is spread at higher predicted earnings.


# Big Picture

Overall, our preliminary analysis indicates that institutions with higher average net price tend to have higher median earnings 10 years after entry, but the relationship is modest and highly variable. Net price alone explains only a small portion of the differences in earnings across colleges, suggesting that other institutional characteristics such as selectivity, student demographics, and program offerings likely play substantial roles.

In practical terms, while there is a positive association between cost and earnings at the institutional level, tuition by itself is not a strong predictor of long-term financial outcomes. In the next stages of the project, we plan to extend this model by incorporating additional covariates to better isolate and understand the factors associated with variation in median earnings across institutions.
