---
title: "Stat 330 Final Project"
format: html
editor: visual
---

## Stat 330 Final Project

```{r setup, include=FALSE}
library(tidyverse)
```

```{r}
#Read your data into R. 
proj_data <- read.csv("college-scorecard-institution.csv")
```

**Note:** Loads the College Scorecard dataset. All analysis below is based on this dataset.

```{r}
# Variables to clean and convert to numeric
vars_to_clean <- c(
  "NPT4_PUB", "NPT4_PRIV", "MD_EARN_WNE_P10",
  "PCTPELL", "CONTROL", "SATVRMID", "SATMTMID",
  "ADM_RATE", "UGDS"
)

proj_data[vars_to_clean] <- lapply(proj_data[vars_to_clean], function(x) {
  x[x == "NULL"] <- NA
  as.numeric(x)
})

# Create one price variable for public and private schools
proj_data$NET_PRICE <- ifelse(!is.na(proj_data$NPT4_PUB),
                              proj_data$NPT4_PUB,
                              proj_data$NPT4_PRIV)

# Student quality
proj_data$SAT_AVG <- (proj_data$SATVRMID + proj_data$SATMTMID) / 2

# Log earnings
proj_data$LOG_EARN <- log(proj_data$MD_EARN_WNE_P10)

# fix negative net price
proj_data$NET_PRICE[proj_data$NET_PRICE < 0] <- NA

summary(proj_data[, c("LOG_EARN", "NET_PRICE", "PCTPELL", "SAT_AVG", "ADM_RATE", "UGDS", "CONTROL")])

nrow(na.omit(proj_data[, c("LOG_EARN", "NET_PRICE", "PCTPELL", "SAT_AVG", "ADM_RATE", "UGDS", "CONTROL")]))
```

**Note:** Cleans key variables by converting "NULL" values to NA and then converting them to numeric. Also creates: - `NET_PRICE` (combined public/private cost) - `SAT_AVG` (proxy for student quality) - `LOG_EARN` (log-transformed earnings) Negative net price values are removed since they are not meaningful.

The summary output checks variable ranges, and the row count shows how many complete observations remain for the full model.

```{r}
nrow(na.omit(proj_data[, c("LOG_EARN", "NET_PRICE", "PCTPELL")]))
nrow(na.omit(proj_data[, c("LOG_EARN", "NET_PRICE", "PCTPELL", "SAT_AVG")]))
```

**Note:** Compares how many observations are available under simpler vs more complex models. Adding variables like SAT reduces sample size due to missing data.

```{r}
# EDA 1
ggplot(proj_data, aes(x = MD_EARN_WNE_P10)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white", na.rm = TRUE) +
  labs(x = "Median Earnings (USD)", y = "Count") +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(labels = scales::dollar)
```

**Note:** Shows the distribution of median earnings. The right-skewed shape motivates using a log transformation in later models.

```{r}
# EDA 2
ggplot(proj_data, aes(x = NET_PRICE, y = MD_EARN_WNE_P10)) +
  geom_point(alpha = 0.4, na.rm = TRUE) +
  labs(title = "Price vs Median Earnings",
       x = "Average Tuition Price",
       y = "Median Earnings") +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(labels = scales::comma)
```

**Note:** Displays the raw relationship between cost and earnings before controlling for other variables.

```{r}
ggplot(proj_data, aes(x = PCTPELL, y = MD_EARN_WNE_P10)) +
  geom_point(alpha = 0.4, na.rm = TRUE) +
  labs(title = "Percent Pell Grant Recipients vs Median Earnings",
       x = "Percent of Students Receiving Pell Grants",
       y = "Median Earnings") +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(labels = scales::comma)
```

**Note:** Shows the relationship between socioeconomic composition (Pell share) and earnings.

```{r}
lm.fit <- lm(MD_EARN_WNE_P10 ~ NET_PRICE, data = proj_data)
summary(lm.fit)
```

**Note:** Baseline model from the proposal using only net price to predict earnings. Used as a reference point for later models.

```{r}
# Residuals vs fitted
plot(lm.fit$fitted.values,
     resid(lm.fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red")

# Q-Q plot
qqnorm(resid(lm.fit),
       main = "Normal Q-Q Plot of Residuals")
qqline(resid(lm.fit), col = "red")
```

**Note:** Diagnostic plots for the baseline model. Suggests potential issues (non-normality/skewness), motivating use of log earnings.

```{r}
plot(proj_data$NET_PRICE, proj_data$LOG_EARN,
     xlab = "Average Net Price ($)",
     ylab = "log(Median Earnings)",
     pch = 16, cex = 0.5,
     xaxt = "n",
     col = c("steelblue", "darkorange", "forestgreen")[proj_data$CONTROL])
axis(1, at = seq(0, 100000, by = 25000),
     labels = c("$0", "$25k", "$50k", "$75k", "$100k"))
legend("bottomright",
       legend = c("Public", "Private NP", "Private FP"),
       col = c("steelblue", "darkorange", "forestgreen"),
       pch = 16, cex = 0.7)
```

**Note:** Examines the relationship on the log scale and introduces institution type through color coding.

```{r}
plot(proj_data$PCTPELL, proj_data$LOG_EARN,
     xlab = "Proportion Receiving Pell Grants",
     ylab = "log(Median Earnings)",
     main = "Pell Grant % vs. log(Earnings)",
     pch = 16, cex = 0.5,
     col = c("steelblue", "darkorange", "forestgreen")[proj_data$CONTROL])

legend("topright",
       legend = c("Public", "Private NP", "Private FP"),
       col = c("steelblue", "darkorange", "forestgreen"),
       pch = 16, cex = 0.7)
```

**Note:** Shows how the Pell relationship behaves after log transformation and across institution types.

## Model Development and Comparison

```{r}
mod1 <- lm(LOG_EARN ~ NET_PRICE + PCTPELL, data = proj_data)
summary(mod1)

mod2 <- lm(LOG_EARN ~ NET_PRICE + PCTPELL + factor(CONTROL), data = proj_data)
summary(mod2)

mod3 <- lm(LOG_EARN ~ NET_PRICE * factor(CONTROL) + PCTPELL, data = proj_data)
summary(mod3)

AIC(mod1, mod2, mod3)


# Model comparison
summary(mod1)$adj.r.squared
summary(mod2)$adj.r.squared
summary(mod3)$adj.r.squared
```

**Note:** These models progressively add complexity: - mod1: basic model (cost + Pell) - mod2: adds institution type - mod3: adds interaction between cost and institution type

Model fit improves as complexity increases, suggesting institutional differences matter.

```{r}
mod4 <- lm(LOG_EARN ~ NET_PRICE * factor(CONTROL) +
                         PCTPELL + SAT_AVG + ADM_RATE + UGDS,
           data = proj_data)

summary(mod4)

proj_data$LOG_PRICE <- log(proj_data$NET_PRICE)

mod5 <- lm(LOG_EARN ~ LOG_PRICE * factor(CONTROL) +
                         PCTPELL + SAT_AVG + ADM_RATE + UGDS,
           data = proj_data)

summary(mod5)

AIC(mod4, mod5)
```

**Note:** These models add additional controls such as SAT (student quality), admission rate, and enrollment. These help isolate whether cost still matters after accounting for student characteristics.

mod5 also tests a log transformation of price.

```{r}
# final model
mod_final <- lm(LOG_EARN ~ NET_PRICE +
                              PCTPELL +
                              SAT_AVG +
                              UGDS +
                              factor(CONTROL),
                data = proj_data)

summary(mod_final)
AIC(mod_final)
```

**Note:** This is the final selected model. It balances interpretability and performance by keeping key predictors (cost, Pell share, SAT, size, and institution type) while removing unnecessary complexity.

```{r}
nrow(na.omit(proj_data[, c("LOG_EARN", "NET_PRICE", "PCTPELL")]))
nrow(na.omit(proj_data[, c("LOG_EARN", "NET_PRICE", "PCTPELL", "SAT_AVG", "ADM_RATE", "UGDS", "CONTROL")]))
```

```{r}
summary(proj_data[, c("MD_EARN_WNE_P10", "NET_PRICE", "PCTPELL", "UGDS", "SAT_AVG")])
```

**Note:** Using this for the Data Description part.

```{r}
library(caret)

# set seed for reproducibility
set.seed(330)

# remove rows with missing values for final model variables
cv_data <- na.omit(proj_data[, c("LOG_EARN", "NET_PRICE", "PCTPELL", 
                                  "SAT_AVG", "UGDS", "CONTROL")])
cv_data$CONTROL <- factor(cv_data$CONTROL)

# 10-fold CV setup
ctrl <- trainControl(method = "cv", number = 10)

# mod1: basic model
cv_mod1 <- train(LOG_EARN ~ NET_PRICE + PCTPELL,
                 data = cv_data,
                 method = "lm",
                 trControl = ctrl)

# mod2: adds institution type
cv_mod2 <- train(LOG_EARN ~ NET_PRICE + PCTPELL + CONTROL,
                 data = cv_data,
                 method = "lm",
                 trControl = ctrl)

# final model: full
cv_final <- train(LOG_EARN ~ NET_PRICE + PCTPELL + SAT_AVG + UGDS + CONTROL,
                  data = cv_data,
                  method = "lm",
                  trControl = ctrl)

# compare results
results <- resamples(list(Basic = cv_mod1,
                          WithType = cv_mod2,
                          Final = cv_final))
summary(results)
```

```{r}
# Figure 1 - earnings histogram
pdf("figure1_earnings.pdf", width = 6, height = 4)
ggplot(proj_data, aes(x = MD_EARN_WNE_P10)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white", na.rm = TRUE) +
  labs(x = "Median Earnings (USD)", y = "Count") +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(labels = scales::dollar)
dev.off()

# Figure 2 - net price vs log earnings by institution type
pdf("figure2_netprice.pdf", width = 6, height = 4)
plot(proj_data$NET_PRICE, proj_data$LOG_EARN,
     xlab = "Average Net Price ($)",
     ylab = "log(Median Earnings)",
     pch = 16, cex = 0.5,
     xaxt = "n",
     col = c("steelblue", "darkorange", "forestgreen")[proj_data$CONTROL])
axis(1, at = seq(0, 100000, by = 25000),
     labels = c("$0", "$25k", "$50k", "$75k", "$100k"))
legend("bottomright",
       legend = c("Public", "Private NP", "Private FP"),
       col = c("steelblue", "darkorange", "forestgreen"),
       pch = 16, cex = 0.7)
dev.off()

# Figure 3
pdf("figure3_pellgrant.pdf", width = 6, height = 4)
plot(proj_data$PCTPELL, proj_data$LOG_EARN,
     xlab = "Proportion Receiving Pell Grants",
     ylab = "log(Median Earnings)",
     main = "Pell Grant % vs. log(Earnings)",
     pch = 16, cex = 0.5,
     col = c("steelblue", "darkorange", "forestgreen")[proj_data$CONTROL])

legend("topright",
       legend = c("Public", "Private NP", "Private FP"),
       col = c("steelblue", "darkorange", "forestgreen"),
       pch = 16, cex = 0.7)
dev.off()
```
